---
title: "Epigenomics Project R code"
author: "Paola Maragno - Carlo Manenti"
date: "2023-02-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



#EPIGENOMICS PROJECT (R - CODE)



## Getting started 
We need to set the working directory
```{r}
setwd("~/Downloads/data&analysis")
```



## PEAKs BOX PLOTs 
write somenthing funny



### rep 1 - encode (peak overlap)
```{r}
rep1_encode_p <- as.data.frame(read.table("commonpeaks_rep1_encode.bed",header = FALSE, sep="\t",stringsAsFactors=FALSE))
rep1_nonen_p <- as.data.frame(read.table("peaks_rep1_nonin_enc.bed",header = FALSE, sep="\t",stringsAsFactors=FALSE))

boxplot(10^-rep1_encode_p$V9, 10^-rep1_nonen_p$V9,ylim=c(0,10^-5),names = c('overlapping with ENCODE','non overlapping with ENCODE'),       
        ylab ='q_value',main='q_value of peaks of replicate 1 overlapping or not with ENCODE peaks - overlap')

```



### rep 2 - encode (peak overlap)
```{r}
rep2_encode_p <- as.data.frame(read.table("commonpeaks_rep2_encode.bed",header = FALSE, sep="\t",stringsAsFactors=FALSE))
rep2_nonen_p <- as.data.frame(read.table("peaks_rep2_nonin_enc.bed",header = FALSE, sep="\t",stringsAsFactors=FALSE))

boxplot(10^-rep2_encode_p$V9, 10^-rep2_nonen_p$V9,ylim=c(0,10^-5),names = c('overlapping with ENCODE','non overlapping with ENCODE'),       
        ylab ='q_value',main='q_value of peaks of replicate 2 overlapping or not with ENCODE peaks - overlap')
```



### common rep 1-2 - encode (peak overlap) 
#DOES NOT WORK AS INTENEDED
```{r}
com_rep_encode_p <- as.data.frame(read.table("commonpeaks_encode.bed",header = FALSE, sep="\t",stringsAsFactors=FALSE))
com_rep_nonencode_p <- as.data.frame(read.table("rep_commonpeaks_nonin_encode.bed",header = FALSE, sep="\t",stringsAsFactors=FALSE))

boxplot(10^-com_rep_encode_p$V5, 10^-com_rep_nonencode_p$V5,ylim=c(0,10^-7.5),names = c('overlapping with ENCODE','non overlapping with ENCODE'),       
        ylab ='q_value',main='q_value of common peaks between replicate 1-2 overlapping or not with ENCODE peaks - overlap')

```



### merged - encode (peak overlap)
```{r}
merg_encode_p <- as.data.frame(read.table("mergedpeaks_encode.bed",header = FALSE, sep="\t",stringsAsFactors=FALSE))
merg_nonencode_p <- as.data.frame(read.table("peaks_merged_nonin_encode.bed",header = FALSE, sep="\t",stringsAsFactors=FALSE))

boxplot(10^-merg_encode_p$V9, 10^-merg_nonencode_p$V9,ylim=c(0,10^-5),names = c('overlapping with ENCODE','non overlapping with ENCODE'),       
        ylab ='q_value',main='q_value of peaks from merged replicates overlapping or not with ENCODE peaks - overlap')
```



## SUMMITs BOX PLOTs
something SillY



### rep 1 - encode (summit proximity)
```{r}
rep1_encode_s <- as.data.frame(read.table("summit_rep1_in_encode.bed",header = FALSE, sep="\t",stringsAsFactors=FALSE))
rep1_nonen_s <- as.data.frame(read.table("summit_rep1_nonin_enc.bed",header = FALSE, sep="\t",stringsAsFactors=FALSE))

boxplot(10^-rep1_encode_s$V5, 10^-rep1_nonen_s$V5,ylim=c(0,10^-5),names = c('overlapping with ENCODE','non overlapping with ENCODE'),       
        ylab ='p_value',main='p_value of peaks of replicate 1 overlapping or not with ENCODE peaks - summit proximity')
```



### rep 2 - encode (summit proximity)
```{r}
rep2_encode_s <- as.data.frame(read.table("summit_rep2_in_encode.bed",header = FALSE, sep="\t",stringsAsFactors=FALSE))
rep2_nonen_s <- as.data.frame(read.table("summit_rep2_nonin_enc.bed",header = FALSE, sep="\t",stringsAsFactors=FALSE))

boxplot(10^-rep2_encode_s$V5, 10^-rep2_nonen_s$V5,ylim=c(0,10^-5),names = c('overlapping with ENCODE','non overlapping with ENCODE'),       
        ylab ='p_value',main='p_value of peaks of replicate 2 overlapping or not with ENCODE peaks - summit proximity')
```



### common rep 1-2 - encode (summit proximity)
```{r}
com_rep_encode_s <- as.data.frame(read.table("commonsummit_in_encode.bed",header = FALSE, sep="\t",stringsAsFactors=FALSE))
com_rep_nonencode_s <- as.data.frame(read.table("rep_commonsummit_nonin_encode.bed",header = FALSE, sep="\t",stringsAsFactors=FALSE))

boxplot(10^-com_rep_encode_s$V5, 10^-com_rep_nonencode_s$V5,ylim=c(0,10^-6.5),names = c('overlapping with ENCODE','non overlapping with ENCODE'),       
        ylab ='p_value',main='p_value of common peaks between replicate 1-2 overlapping or not with ENCODE peaks - summit proximity')
```



### merged - encode (summit proximity)
```{r}
merg_encode_s <- as.data.frame(read.table("summit_merged_in_encode.bed",header = FALSE, sep="\t",stringsAsFactors=FALSE))
merg_nonencode_s <- as.data.frame(read.table("summit_merged_nonin_encode.bed",header = FALSE, sep="\t",stringsAsFactors=FALSE))

boxplot(10^-merg_encode_s$V5, 10^-merg_nonencode_s$V5,ylim=c(0,10^-5),names = c('overlapping with ENCODE','non overlapping with ENCODE'),       
        ylab ='p_value',main='p_value of peaks from merged replicates overlapping or not with ENCODE peaks - summit proximity')
```



## CHROMATIN STATEs
...


retrieve the coordinates of the chromatin states each one with the number of summits falling in it
```{r}
chromatin <- as.data.frame(read.table("summits_per_chromatin_state.bed",header = FALSE, sep="\t",stringsAsFactors=FALSE))
```

# retrieve the coordinates of the summits
```{r}
summits <- as.data.frame(read.table("chr_state_each_summit.bed",header = FALSE, sep="\t",stringsAsFactors=FALSE))

# initialize a data frame in which saving the number of summits falling in each chromatin state
chr_states <- data.frame()

# list of the 18 chromatin states
chr_states_names <- c('1_TssA','2_TssFlnk','3_TssFlnkU','4_TssFlnkD','5_Tx','6_TxWk','7_EnhG1',
                      '8_EnhG2','9_EnhA1','10_EnhA2','11_EnhWk','12_ZNF/Rpts','13_Het','14_TssBiv',
                      '15_EnhBiv','16_ReprPC','17_ReprPCWk','18_Quies')
 
for (state in chr_states_names) {
  state_table <- chromatin[which(chromatin$V4 == state),]
  total_summits <- sum(state_table$V10)
  chr_states <- rbind(chr_states, c(state,total_summits))
}
colnames(chr_states) <- c('Chromatin_state','Overlapping_summits')

# compute the number of summits not falling in any of the reported chromatin state and add this information to the data frame
summits_not_overlapping <- nrow(summits) - sum(as.numeric(chr_states$Overlapping_summits))
chr_states <- rbind(chr_states,c('other states',summits_not_overlapping))
chr_states$overlapping_summits<-as.numeric(chr_states$Overlapping_summits)

# plotting the informations
library(ggplot2)

chr_states$Chromatin_state <- factor(chr_states$Chromatin_state, levels = chr_states$Chromatin_state)
ggplot(chr_states,aes(x = Chromatin_state, y=overlapping_summits))+ 
  geom_bar(stat="identity")+ 
  theme_classic() +
  ylim(0,max(chr_states$overlapping_summits)+10)+
  labs(title='Number of summits falling within each chromatin state')

```

## GREAT 
...



### ALL GENES
```{r}
target_genes <- as.data.frame(read.table("target_genes.txt",header = FALSE, sep="\t",stringsAsFactors=FALSE))
nrow(target_genes)
# number of regulated genes: 3282

library(stringr)
t <- str_extract_all(target_genes$V2, "\\w+")
 
r1 <- vapply(t,
       FUN=function(x){x[2]},
       FUN.VALUE=character(1)
)
r2 <- vapply(t,
             FUN=function(x){x[4]},
             FUN.VALUE=character(1)
)

r2 <- r2[!is.na(r2)]
tot <- as.numeric(c(r1,r2))
length(tot[tot <= 1000])
# 472 GATA2 binds the promoter

length(tot[tot > 1000])
# 3400 GATA2 bind s distal regulatory region

'GATA2' %in% target_genes$V1
# GATA2 regulates itself 
```



### TOP 5K GENES
```{r}
target_genes_5000 <- as.data.frame(read.table("target_genes5000.txt",header = FALSE, sep="\t",stringsAsFactors=FALSE))
nrow(target_genes_5000)
# number of regulated genes: 2627

library(stringr)
t <- str_extract_all(target_genes_5000$V2, "\\w+")

r1 <- vapply(t,
             FUN=function(x){x[2]},
             FUN.VALUE=character(1)
)
r2 <- vapply(t,
             FUN=function(x){x[4]},
             FUN.VALUE=character(1)
)

r2 <- r2[!is.na(r2)]
tot <- as.numeric(c(r1,r2))
length(tot[tot <= 1000])
# 348 GATA2 binds the promoter

length(tot[tot > 1000])
# 2680 GATA2 bind s distal regulatory region

'GATA2' %in% target_genes_5000$V1
# Same as before
```



##DNAse-seq 
...



```{r}
ov <- 77.757
nnov <- 100 - ov
nnov

library(RColorBrewer)
myPalette <- brewer.pal(3, "Set2") 
labels <- c('overlapping - 77.757%','non overlapping - 22.243%')
pie(c(ov,nnov),labels,col=myPalette,border='white',main = 'Percentage of GATA2 summits overlapping with 
    DNAse-seq regions')
```

# a quick try to define the chromatin states of summits non overlapping with DNase
```{r}
summits <- as.data.frame(read.table("chr_state_each_summit_non_in_DNAse.bed",header = FALSE, sep="\t",stringsAsFactors=FALSE))

# list of the 18 chromatin states
chr_states_names <- c('1_TssA','2_TssFlnk','3_TssFlnkU','4_TssFlnkD','5_Tx','6_TxWk','7_EnhG1',
                      '8_EnhG2','9_EnhA1','10_EnhA2','11_EnhWk','12_ZNF/Rpts','13_Het','14_TssBiv',
                      '15_EnhBiv','16_ReprPC','17_ReprPCWk','18_Quies')

# initialize a data frame in which saving the number of summits falling in each chromatin state
chr_states <- data.frame()
 
for (state in chr_states_names) {
  state_n <- nrow(summits[which(summits$V9 == state),])
  chr_states <- rbind(chr_states, c(state, state_n))
}
colnames(chr_states) <- c('Chromatin_state', 'Overlapping_summits')

chr_states$Overlapping_summits <- as.numeric(chr_states$Overlapping_summits)
#rownames(chr_states) <- chr_states_names

# compute the number of summits not falling in any of the reported chromatin state and add this information to the data frame
summits_not_overlapping <- nrow(summits) - sum(chr_states$Overlapping_summits)
chr_states <- rbind(chr_states, c('other states', summits_not_overlapping))

chr_states$Overlapping_summits <- as.numeric(chr_states$Overlapping_summits)

# plotting the informations
library(ggplot2)

chr_states$Chromatin_state <- factor(chr_states$Chromatin_state, levels = chr_states$Chromatin_state)

ggplot(chr_states,aes(x = Chromatin_state, y=Overlapping_summits))+ 
  geom_bar(stat="identity")+ 
  theme_classic() +
  ylim(0,max(chr_states$Overlapping_summits)+10)+
  labs(title='Number of summits falling within each chromatin state')

```


